---
title: "Untitled"
author: "rachel_cassway"
date: "2025-04-02"
output: html_document
---

### Load and Clean Data
Getting data into the following format:
- Rows = users
- Columns = recipes
- Values = ratings (stars)

```{r}
# load libraries
library(dplyr)
library(tidyr)
library(proxy)
library(scales)

# load csv into dataframe
df <- read.csv("recipe_review_data.csv")

# select collab filtering columns
df <- df %>%
  mutate(
    user_id = trimws(as.character(user_id)),
    recipe_code = trimws(as.character(recipe_code)),
    stars = as.numeric(stars)
  )

# clean ratings: take average if user rated the same item multiple times
ratings_cleaned <- df %>%
  filter(!is.na(stars)) %>%
  group_by(user_id, recipe_code) %>%
  summarise(stars = mean(stars, na.rm = TRUE), .groups = "drop")

# pivot_wider, to prep for matrix
rating_wide <- ratings_cleaned %>%
  pivot_wider(names_from = recipe_code, values_from = stars)

# convert to data.frame before setting rownames
rating_wide <- as.data.frame(rating_wide)

# set user_id as rownames
rating_wide$user_id <- trimws(as.character(rating_wide$user_id))
rownames(rating_wide) <- rating_wide$user_id
rating_wide <- rating_wide %>% select(-user_id)

# convert to numeric matrix
rating_wide[] <- lapply(rating_wide, function(x) as.numeric(as.character(x)))
rating_matrix <- as.matrix(rating_wide)

# preview matrix
head(rating_matrix)

# make sure each recipe is distinct in lookup table
recipe_lookup <- df %>%
  select(recipe_code, recipe_name) %>%
  distinct(recipe_code, .keep_all = TRUE)  

```
### Some analysis on data to explore popular recipes and frequent users
```{r}
# which users have rated the most recipes
user_rating_counts <- rowSums(!is.na(rating_matrix))
top_users <- sort(user_rating_counts, decreasing = TRUE)

head(top_users)

length(user_rating_counts[user_rating_counts >= 10])

# which recipe has the most ratings
recipe_popularity <- df %>%
  filter(!is.na(stars)) %>%
  count(recipe_code, recipe_name, sort = TRUE)

head(recipe_popularity)

# users who rated most recipes, with highest user reputation
top_reviewers <- df %>%
  filter(!is.na(stars)) %>%
  group_by(user_id, user_name) %>%
  summarise(
    num_reviews = n(),
    avg_rating = mean(stars),
    user_reputation = max(user_reputation),
    .groups = "drop" 
  ) %>%
  arrange(desc(num_reviews))


```
### user-user collaborative filtering function
```{r}
user_user_cf <- function(df, user, sim_metric = "Cosine", k = 2, 
                         recipe_lookup_df = NULL) {
  
  # if user not in data, return not found message
  if (!(user %in% rownames(df))) {
    message("Error: User not found in dataset.")
    return(NULL)
  }

  # if the user has no missing values, return message
  if (all(!is.na(df[user, ]))) {
    message("User has no missing ratings to predict.")
    return(NULL)
  }

  # if similarity metric not Cosine or L2, return message
  if (!(sim_metric %in% c("Cosine", "L2"))) {
    message("Unsupported similarity metric. Use 'Cosine' or 'L2'.")
    return(NULL)
  }

  # 1. center data by row means
  user_means <- rowMeans(df, na.rm = TRUE)
  centered_data <- sweep(df, 1, user_means, FUN = "-")
  # fill NaN values with 0 for user-user only
  centered_data[is.na(centered_data)] <- 0  

  # 2. get target user vector
  user_vector <- as.matrix(centered_data[user, , drop = FALSE])

  # 3. compute similarity
  if (sim_metric == "Cosine") {
    sim_vec <- proxy::simil(user_vector, centered_data, method = "cosine")
    similarity_scores <- as.vector(sim_vec)
  } else if (sim_metric == "L2") {
    diff <- centered_data - matrix(rep(user_vector, nrow(centered_data)), 
                                   nrow = nrow(centered_data), byrow = TRUE)
    similarity_scores <- -sqrt(rowSums(diff^2))
  }

  # 4. normalize similarity
  similarity_scores <- rescale(similarity_scores, to = c(0, 1))
  sim_df <- data.frame(User = rownames(df), Similarity = similarity_scores)
  sim_df <- sim_df %>% filter(User != user)

  # 5. select top k similar users
  top_k_users <- sim_df %>%
  arrange(desc(Similarity)) %>%
  slice_head(n = min(k, nrow(sim_df)))


  # 6. prepare ratings for predictions
  k_ratings <- df[top_k_users$User, ]
  k_ratings <- t(apply(k_ratings, 1, function(row) {
    row[is.na(row)] <- mean(row, na.rm = TRUE)
    return(row)
  }))

  # 7. weighted prediction
  if (sum(top_k_users$Similarity) == 0) {
    message("Warning: All similarities are zero. Returning NULL.")
    return(NULL)
  }

  # calc ratings
  weighted <- sweep(k_ratings, 1, top_k_users$Similarity, FUN = "*")
  predicted_ratings <- colSums(weighted) / sum(top_k_users$Similarity)

  # return missing predictions only for target user
  result <- data.frame(recipe_code = colnames(df), Predicted_Rating = predicted_ratings)
  missing_items <- is.na(df[user, ])
  result <- result[missing_items, ]
  
  # add recipe names if provided in original df
  if (!is.null(recipe_lookup_df)) {
    result <- result %>%
      left_join(recipe_lookup_df, by = "recipe_code") %>%
      select(recipe_code, recipe_name, Predicted_Rating)
  }

  result <- result %>%
    arrange(desc(Predicted_Rating))
  
  return(result)
}

```



### item-item collaborative filtering function
```{r}
item_item_cf <- function(df, user, sim_metric = "Cosine", k = 2, 
                         recipe_lookup_df = NULL) {
  
  # if user not in data, return not found message
  if (!(user %in% rownames(df))) {
    message("Error: User not found in dataset.")
    return(NULL)
  }

  # if the user has no missing values, return message
  if (all(!is.na(df[user, ]))) {
    message("User has no missing ratings to predict.")
    return(NULL)
  }

  # if similarity metric not Cosine or L2, return message
  if (!(sim_metric %in% c("Cosine", "L2"))) {
    message("Unsupported similarity metric. Use 'Cosine' or 'L2'.")
    return(NULL)
  }
  
  # 1. center by column (item) means
  item_means <- colMeans(df, na.rm = TRUE)
  centered_data <- sweep(df, 2, item_means, FUN = "-")
  centered_data[is.na(centered_data)] <- 0

  # 2. transpose for item-based similarity (now rows = items)
  item_data <- t(centered_data)

  # 3. ratings the user has already made
  user_vector <- as.numeric(df[user, ])
  names(user_vector) <- colnames(df)

  predictions <- numeric(ncol(df))
  names(predictions) <- colnames(df)

  for (item in names(user_vector)) {
    # skip if user already rated it
    if (!is.na(user_vector[item])) next

    # get all items the user has rated
    rated_items <- names(user_vector)[!is.na(user_vector)]

    # compute similarity between this target item and all items the user rated
    if (sim_metric == "Cosine") {
      sim_scores <- proxy::simil(as.matrix(item_data[item, , drop = FALSE]),
                                 item_data[rated_items, , drop = FALSE],
                                 method = "cosine")
      sim_scores <- as.vector(sim_scores)
    } else if (sim_metric == "L2") {
      diff <- item_data[item, ] - item_data[rated_items, ]
      sim_scores <- -sqrt(rowSums(diff^2))
    }

    # 4. normalize similarity
    sim_scores <- rescale(sim_scores, to = c(0, 1))

    # grab the user’s ratings for the similar items
    rated_scores <- user_vector[rated_items]

    # 5. select top k similar users
    top_k <- order(sim_scores, decreasing = TRUE)[1:min(k, length(sim_scores))]
    top_k_sims <- sim_scores[top_k]
    top_k_sims <- top_k_sims[!is.na(top_k_sims)]
    top_k_ratings <- rated_scores[top_k]

    # 6. make predictions
    if (sum(top_k_sims) == 0) {
      predictions[item] <- NA
    } else {
      predictions[item] <- sum(top_k_sims * top_k_ratings) / sum(top_k_sims)
    }
  }

  # convert predictions to a data frame
  result <- data.frame(recipe_code = names(predictions), Predicted_Rating = predictions)
  result <- result[is.na(df[user, ]), ]

  # add recipe names if provided
  if (!is.null(recipe_lookup_df)) {
    result <- result %>%
      left_join(recipe_lookup_df, by = "recipe_code") %>%
      select(recipe_code, recipe_name, Predicted_Rating)
  }

  result <- result %>%
    filter(!is.na(Predicted_Rating)) %>%
    arrange(desc(Predicted_Rating))

  return(result)
}

```


### helper function to find rating for one target user and one target recipe
```{r}
predict_one_recipe <- function(rating_matrix, user_id, recipe_code, k = 5, 
                               sim_metric = "Cosine", method = "user", 
                               recipe_lookup_df = NULL) {
  # Check method type
  if (!(method %in% c("user", "item"))) {
    stop("Method must be either 'user' or 'item'")
  }

  if (method == "user") {
    # run user-user collaborative filtering
    predictions <- user_user_cf(
      df = rating_matrix,
      user = user_id,
      sim_metric = sim_metric,
      k = k,
      recipe_lookup_df = recipe_lookup_df
    )

  } else if (method == "item") {
    # run item-item collaborative filtering 
    predictions <- item_item_cf(
      df = rating_matrix,
      user = user_id,
      sim_metric = sim_metric,
      k = k,
      recipe_lookup_df = recipe_lookup_df
    )

  }
  
  # filter to just the target recipe
  prediction <- predictions[predictions$recipe_code == recipe_code, ]

  return(prediction)
}

```

### Testing Collaborative Filtering Functions
```{r}
# find users with at least one missing rating to predict for
users_with_missing <- rownames(rating_matrix)[
  apply(rating_matrix, 1, function(row) any(is.na(row)))
]

# test user
test_user <- users_with_missing[50]
print(paste("Testing user:", test_user))

# predict for chosen test user with user-user CF
predicted_ratings <- user_user_cf(rating_matrix, user = test_user, 
                                  sim_metric = "Cosine", k = 5, 
                                  recipe_lookup_df = recipe_lookup)

# display predictions
print(paste("USER-USER: Predicted ratings for missing items for test user:", test_user))
head(predicted_ratings)

# predict for chosen test user with item-item CF
predicted_ratings <- item_item_cf(rating_matrix, user = test_user, 
                                  sim_metric = "Cosine", k = 5, 
                                  recipe_lookup_df = recipe_lookup)

# display predictions
print(paste("ITEM-ITEM: Predicted ratings for missing items for test user:", test_user))
head(predicted_ratings)
```

```{r}
# find rating for target user and target recipe 
target_user <- "u_05PZUpOV27Pv"
target_recipe <- "3309"
head(recipe_popularity)

predict_one_recipe(rating_matrix, target_user, target_recipe, k = 5, method="user", 
                   recipe_lookup_df = recipe_lookup)

predict_one_recipe(rating_matrix, target_user, target_recipe, k = 5, method="item", 
                   recipe_lookup_df = recipe_lookup)
```

### Testing CF Models by Hiding Ratings
In order to test the capabilities the models, we will 
1. pick a user with a good number of ratings
2. randomly hide a few of their actual ratings (setting to NA)
3. run prediction for those recipes
4. compare predicted vs actual ratings


```{r}

evaluate_user_cf <- function(test_user, rating_matrix, seed = 42, k = 5, 
                             n_holdout = 5, sim_metric = "Cosine", 
                             recipe_lookup_df = NULL) {
  # find the items this user has rated
  rated_items <- which(!is.na(rating_matrix[test_user, ]))
  
  # randomly hold out 5 of their ratings - 
  set.seed(seed)  # for reproducibility
  held_out_items <- sample(rated_items, n_holdout)
  
  # store true ratings to compare later
  true_ratings <- rating_matrix[test_user, held_out_items]
  
  # make test matrix (copied matrix with held-out values set to NA)
  test_matrix <- rating_matrix
  test_matrix[test_user, held_out_items] <- NA
  
  # Predict using user-user or item-item
  pred_user_user <- user_user_cf(test_matrix, user = test_user, 
                                 sim_metric = sim_metric, k = k, 
                                 recipe_lookup_df = recipe_lookup_df)
  pred_item_item <- item_item_cf(test_matrix, user = test_user, 
                                 sim_metric = sim_metric, k = k,
                                 recipe_lookup_df = recipe_lookup_df)
  
  # build a full comparison table
  
  # filter both prediction outputs to the held-out recipe codes
  user_preds <- pred_user_user %>%
    filter(recipe_code %in% names(true_ratings)) %>%
    select(recipe_code, recipe_name, user_pred = Predicted_Rating)
  
  item_preds <- pred_item_item %>%
    filter(recipe_code %in% names(true_ratings)) %>%
    select(recipe_code, item_pred = Predicted_Rating)
  
  # merge user and item predictions
  comparison <- user_preds %>%
    left_join(item_preds, by = "recipe_code")
  
  # add actual held-out ratings and user ID
  comparison$actual <- true_ratings[comparison$recipe_code]
  comparison$user_id <- test_user

  # reorder columns for readability
  comparison <- comparison %>%
    select(user_id, recipe_code, recipe_name, user_pred, item_pred, actual)

  
  return(comparison)
}
```



```{r}
# test evaluate function
rated_10 <- user_rating_counts[user_rating_counts >= 10]
test_user <- names(rated_10)[1]  # first user in the group
evaluate_user_cf(test_user, rating_matrix, seed = 42, 
                 k = 5, n_holdout = 5, sim_metric = "Cosine", 
                 recipe_lookup_df = recipe_lookup)
```

### Analyze Models
```{r}
# make matrix for users who rates >= 10 recipes
test_rating_matrix <- rating_matrix[rowSums(!is.na(rating_matrix)) >= 5, ]

# initialize empty data frame
results_df <- data.frame()  

for (user in rownames(test_rating_matrix)) {
  result <- evaluate_user_cf(user, test_rating_matrix, seed = 42, k = 5, 
                             n_holdout = 3, sim_metric = "Cosine", 
                             recipe_lookup_df = recipe_lookup)
  
  if (!is.null(result) && nrow(result) > 0) {
    results_df <- rbind(results_df, result)
  }
}

print(results_df)
```

